\documentclass[margin, 10pt]{res} 
\usepackage{helvet} 

\setlength{\textwidth}{5.1in} 
\usepackage{hyperref}
\hypersetup{colorlinks=true,urlcolor=blue}
\usepackage{xcolor}

\title{Academic CV}

\begin{document}

\moveleft2.5\hoffset\centerline{\large\bf Anna Neufeld} 
 
\moveleft\hoffset\vbox{\hrule width\resumewidth height 1pt}\smallskip 

\begin{resume}

 
\section{Contact Information}
Department of Mathematics and Statistics \hfill  acn2@williams.edu  \\
Williams College \hfill \href{https://anna-neufeld.github.io}{https://anna-neufeld.github.io} \\
18 Hosxey St. \hfill 781-392-4894 \\ Williamstown, MA, 01267 \hfill \\



\section{Employment} 
{\textbf{Assistant Professor of Statistics}} \hfill 2024 - Present \\
Department of Mathematics and Statistics \\
Williams College, Williamstown, MA  \\
\\
{\textbf{Postdoctoral researcher}} \hfill 2023-2024 \\
Biostatistics Program (Public Health Sciences Division) \\
Fred Hutchinson Cancer Center, Seattle, WA
\begin{itemize}
\item Advisor: Dr. Jeffery Leek 
\end{itemize}

\section{Education} 

{\textbf{University of Washington},} Seattle, Washington \hfill 2018-2023 \\
PhD in Statistics
\begin{itemize}
\item Advisor: Dr. Daniela Witten
\item Thesis: Addressing double dipping via selective inference and data thinning 
\end{itemize}

{\textbf{Williams College}}, Williamstown, MA \hfill 2014-2018  \\
BA in Mathematics (Highest Honors) and Computer Science. \textit{Summa Cum Laude} 
\begin{itemize}
\item Advisor: Dr. Brianna Heggeseth
\item Thesis: Regression trees for longitudinal data 
\end{itemize}
 
\section{Teaching Experience}
{\textbf{At Williams College}} 
\begin{itemize}
\item STAT 202, Introduction to Statistical Modeling
\begin{itemize}
\item Fall 2024, 2 sections, 68 total students.
\item Spring 2025, 1 section, 45 students.  
\end{itemize}
\item STAT 442, Statistical Learning and Data Mining 
\begin{itemize}
\item Spring 2025, 12 students.  
\end{itemize}	
\item CSCI 374T, Machine Learning
\begin{itemize}
\item Spring 2021, 10 students.
\item Co-instructed this tutorial with Professor Andrea Danyluk, while serving as a remote visiting lecturer in the Department of Computer Science. 
\end{itemize}
\end{itemize}
{\textbf{At University of Washington}} 
\begin{itemize}
\item Instructor of Record:
\begin{itemize}
\item Stat 311: Elements of Statistical Methods (Summer 2019). Fully responsible for course of 60 students. 
\item Stat 499C: Undergraduate Research (Autumn 2021-Spring 2023, 6 consecutive quarters). Managed the logistics, enrollment, and course credit associated with the Directed Reading Program. There were 5-12 undergraduate students enrolled for credit each quarter.   
\end{itemize} 
\item Head teaching assistant:
\begin{itemize}
\item Stat 311: Elements of Statistical Methods (Autumn 2019, Winter 2020).  
\item Developed lab assignments, maintained course website, helped write assignments and exams, served as liaison between professor and other TAs. There were $180$ total students in the course. Also led two lab sections of 30 undergraduates each. 
\end{itemize} 
\item Teaching assistant: 
\begin{itemize}
\item Stat 570: Regression Methods for Independent Data (Autumn 2021)
\item Stat 527: Nonparametric Regression (Spring 2021)
\item Stat 311: Elements of Statistical Methods (Autumn 2018, Spring 2020)
\item Stat 423:  Applied Regression and Analysis of Variance (Winter 2019)
\item CSE/Stat 416: Introduction to Machine Learning (Spring 2019) 
\item Responsible for lab sections of around 30 undergraduates each. Also held office hours and graded assignments. Courses listed above range from undergraduate introductory courses (Stat 311 and Stat 416) to core courses for the statistics PhD program (Stat 570). 
\end{itemize}
\end{itemize}


%{\textbf{Undergraduate Teaching Assistant}} \hfill September 2015 - May 2018 \\
%Williams College Departments of Computer Science, Mathematics, and Statistics
%\begin{itemize}
%\item Data Structures and Advanced Programming (Fall 2015), Linear Algebra (Spring 2016, Spring 2017), Abstract Algebra (Fall 2016), Regression and Forecasting (Fall 2017, Spring 2018)
%\item Duties included grading homework, holding office hours, and running review sessions. 
%\end{itemize}
%{\textbf{Peer Tutor and Workshop Leader}} \hfill 2016-2018 \\
%Williams College Office of Academic Resources
%\begin{itemize}
%\item Held one-on-one and drop-in tutoring sessions for microeconomics, macroeconomics, calculus, linear algebra, real analysis, statistics, and computer science.  
%\item Assisted biology research students with data analyses in \texttt{R}. 
%\item Developed and taught a series of workshops in \texttt{R} for students from a variety of departments as part of a pilot for a winter study coding workshop program.  
%\end{itemize}

\section{Preprints} 
\textbf{Anna Neufeld}, Joshua Popp, Lucy L. Gao, Alexis Battle, and Daniela Witten (2023) Negative binomial count splitting for single-cell RNA sequencing data. \\ \href{https://arxiv.org/pdf/2307.12985}{https://arxiv.org/pdf/2307.12985}. \\
\\
Ameer Dharamshi, \textbf{Anna Neufeld}, Lucy L. Gao, Jacob Bien, and Daniela Witten (2024). Decomposing Gaussians with unknown covariance. \emph{Revise and resubmit at Biometrika}. \href{https://arxiv.org/pdf/2409.11497}{https://arxiv.org/pdf/2409.11497} \\
\\
Stephen Salerno, Kentaro Hoffman,  Awan Afiaz, \textbf{Anna Neufeld}, Tyler H. McCormick, and Jeffrey T. Leek (2024) Sample Size Considerations for Post-Prediction Inference. %\\
%\\
%Ameer Dharamshi, \textbf{Anna Neufeld}, Lucy L. Gao, Daniela Witten, and Jacob Bien (2025). Thinning a Wishart Random Matrix. \emph{Submitted to Biometrika}.
\section{Publications} 
Stephen Salerno, Jiacheng Miao, Awan Afiaz, Kentaro Hoffman, \textbf{Anna Neufeld}, Qiongshi Lu, Tyler H. McCormick, and Jeffrey T. Leek (2025) ipd: An R Package for Conducting Inference on Predicted Data. To appear in \emph{Bioinformatics}. \\
\\
\textbf{Anna Neufeld}, Ameer Dharamshi, Lucy L. Gao, Daniela Witten, and Jacob Bien (2024) Discussion of ``data fission: splitting a single data point". \emph{The Journal of the American Statistical Association}. \emph{Invited discussion.}  \\
\\
Ameer Dharamshi, \textbf{Anna Neufeld}, Keshav Motwani, Lucy L. Gao, Daniela Witten, and Jacob Bien (2024) Generalized data thinning using sufficient statistics. \emph{Journal of the American Statistical Association}. \emph{SLDS student paper award awarded to Ameer Dharamshi at JSM, 2024.}  \\
\\
\textbf{Anna Neufeld}, Ameer Dharamshi, Lucy L. Gao, and Daniela Witten (2024) Data thinning for convolution-closed distributions. \emph{Journal of Machine Learning Research}. \\
\\
\textbf{Anna Neufeld}, Lucy L. Gao, Joshua Popp, Alexis Battle, and Daniela Witten (2022) Inference after latent variable estimation for single cell RNA-sequencing data. \emph{Biostatistics}. \emph{Winner of best student paper award at WNAR, 2022.}  \\
\\
\textbf{Anna C. Neufeld}, Lucy L. Gao, and Daniela M. Witten (2022) Tree-Values: selective inference for regression trees. \emph{Journal of Machine Learning Research.}\\
\\
\textbf{Anna Neufeld} and Daniela Witten (2021). Discussion of Breiman’s “Two Cultures”: From Two Cultures to One. \emph{Observational Studies}. \emph{Invited Discussion}.   \\
\\
Maxian, O*., \textbf{Neufeld, A.}*, Talis, E. J.*, Childs, L. M., \& Blackwood, J. C. (2017). Zika virus dynamics: when does sexual transmission matter?. \emph{Epidemics}. \\
\small
(\emph{* denotes equal contribution})
\normalsize

\section{Software} 
\textbf{datathin: splitting a random variable into independent training and test components}. R package available on \href{https://github.com/anna-neufeld/datathin}{github}, with tutorials available on our \href{https://anna-neufeld.github.io/datathin/}{website}.  \\
\\
\textbf{countsplit: splitting integer-valued count matrices.} R package available on \href{https://cran.r-project.org/web/packages/countsplit/index.html}{CRAN}, with tutorials available on our \href{https://anna-neufeld.github.io/countsplit.tutorials/}{website}. \\
\\
\textbf{treevalues: selective inference for regression trees.} R package available on \href{https://github.com/anna-neufeld/treevalues}{github}, with tutorials available on our \href{https://anna-neufeld.github.io/treevalues/}{website}. \\
\\
\textbf{splinetree: longitudinal trees and forests using a spline projection method}. \\
R package available on \href{https://cran.r-project.org/web/packages/splinetree/index.html}{CRAN}. Tutorials available on our \href{https://anna-neufeld.github.io/splinetree/}{website}. 


\section{Invited Presentations} 
\begin{itemize}
\item Statistics Empowering Data Science (SEEDS) Conference, January 2025, University of Southern California, Los Angeles, CA. \emph{Data thinning to avoid double dipping.} 
\item Williams College Science Lunch, October 2024, Williamstown, MA. \emph{What is double dipping and how can we avoid it?}
\item Williams College Statistics Colloquium, September 2024, Williamstown, MA. \emph{What is double dipping and how can we avoid it?}
\item Joint Statistical Meetings, JASA Theory \& Methods Invited Session, Portland, OR. \emph{Discussion of ``Data Fission"}. 
\item ASA Section on Statistical Genetics and Genomics monthly webinar, April 2024, virtual. \item Banff International Research Station Workshop on Statistical Aspects of Trustworthy Machine Learning. February, 2024. Banff, AB. \emph{Data thinning to avoid double dipping}. \href{https://anna-neufeld.github.io/slides/BIRS.pdf}{[slides]}. 
\item University of Washington Department of Genome Sciences and Computational Molecular Biology Program Combi seminar, January 2024, Seattle, WA. \emph{Avoiding double dipping in the analysis of scRNA-seq data}.
\item Channing Division Methods Meeting, Brigham and Women's Hospital and Harvard Medical School, December 2023, virtual. \emph{Avoiding double dipping in the analysis of scRNA-seq data}.
\item Williams College Statistics Colloquium, November 2023, Williamstown, MA. \emph{Data thinning to avoid double dipping}. \href{https://anna-neufeld.github.io/slides/jobtalk.pdf}{[slides]}.
\item Reed College Mathematics Colloquium, October 2023, Portland, OR. \emph{Data thinning to avoid double dipping}
\item North American Machine Learning, Optimization, and Statistics Symposium (NAMOS), June 2023, Vancouver, BC. \emph{Data thinning}. 
\item Private Brands Casual Science Seminar at Amazon, April 2023, virtual. \emph{Data thinning}. \href{https://anna-neufeld.github.io/slides/amazon.pdf}{[slides]}. 
\item Fred Hutchinson Cancer Center, group of Dr. Mike Wu, April 2023, virtual. \emph{Data thinning}. 
\item Williams College Statistics Colloquium, October 2022, virtual. \emph{Avoiding ``double dipping" in the analysis of single cell RNA-sequencing data.} 
\item Joint Statistical Meetings, August 2022. Panelist for: \emph{Leo Breiman's Two Cultures: Introspection, Debate, and Discussion 20 Years Later}. 
\item International Seminar on Selective Inference, June 2022, virtual. \emph{Inference after latent variable estimation for single cell RNA-sequencing data}. \href{https://drive.google.com/file/d/1U4s7xiuz5yOCeDL2hjfIPZPbD7EsmINb/view}{[recording]}, \href{https://drive.google.com/file/d/1KPX8TkT85jHYalxhieyy65PAj18PRbFF/view}{[slides]}.
\item Boston Children's Hospital Vascular Anomalies Group Meeting, June 2022, virtual. \emph{Statistical issues when analyzing single cell RNA-sequencing data}. 
\end{itemize}
\section{Contributed Presentations} 
\begin{itemize}
\item Institute of Mathematical Statistics New Researcher's Conference, August 2024, Corvalis, OR. Speed talk on \emph{data thinning}.
\item Joint Statistical Meetings, topic contributed session on ``Harnessing multiple genomic features to identify disease biomarkers,” August 2024, Portland, OR.  \emph{Avoiding double dipping in the analysis of single-cell RNA sequencing data}. 
\item WNAR 2022 Student Paper Competition Session, June 20202, virtual. \emph{Inference after latent variable estimation for single cell RNA-sequencing data}.
\end{itemize}

\section{Mentoring}
{\textbf{Williams College senior theses}}
\begin{itemize}
\item Matthew Wu, 2025
\end{itemize}
{\textbf{NSF INSPIRE U2 REU, Spelman College}} \hfill Summer 2023 \\
(Increasing Statistical Preparation in Research Education for Underrepresented Undergraduates)
\begin{itemize}
\item Virtual RStudio mentor for two undergraduate researchers; 
Allyanna Lewis and Sasha Villefranche. Both students will be presenting at the Annual Biomedical Research Conference for Minoritized Scientists (ABRCMS). 
\end{itemize}

{\textbf{University of Washington Statistics Directed Reading Program (DRP)}}
\begin{itemize}
  \item Co-founded a program that pairs undergraduates with PhD student mentors for independent studies. Modeled after successful \href{https://sites.google.com/view/drp-network/}{Directed Reading Programs} (DRPs) in mathematics departments at several universities. More information can be found at \href{https://spa-drp.github.io}{our website}. 
  \item Served as the graduate student coordinator from 2020-2023. Managed admissions, recruitment, and scheduling for 95 undergraduate independent studies. Several undergraduates have gone on to graduate school in statistics or related fields. 
  \item Served as the instructor of record for Stat 499 (the associated course that undergraduate participants can register for) from 2021-2023. 
  \item Served as a mentor for the following eight undergraduates:
  \begin{itemize}
  \item Winter 2020, Christina Nick, Statistical Natural Language Processing
  \item Spring 2020, Rachael Ren, Infectious Disease Modeling with Differential Equations. See \href{https://spa-drp.github.io/writeups/spring2020/rachael.pdf}{project writeup} and \href{https://spa-drp.github.io/writeups/spring2020/rachaelslides.pdf}{presentation}. Now a PhD student in statistics at the University of Texas. 
  \item Autumn 2020, Harper Zhu, Infectious Disease Modeling on a Network.  See \href{https://spa-drp.github.io/writeups/aut2020/harper-slides.pdf}{presentation} and \href{https://harperzhu.shinyapps.io/DiseaseSimulation/}{shiny app}. Now a MS studnent in data science at Cornell Tech. 
  \item Spring 2021, Kayla Kenyon, Infectious Disease Modeling. See \href{https://spa-drp.github.io//writeups/spring2021/kayla_slides.pdf}{presentation}. 
  \item Autumn 2021, Cathy Qi, Multiple Testing. 
  \item Winter 2022, Hisham Bhatti, Statistical Concepts in Clinical Trials.
 \item Spring 2022, Wei Jun Tan \href{https://spa-drp.github.io/writeups/spring2022/slides/weijun.pdf}{[presentation]} and Iris Zhao \href{https://spa-drp.github.io//writeups/spring2022/slides/iris.pdf}{[presentation]}, Introduction to Computational Biology, co-mentored with Alan Min.  
 \end{itemize}
 \end{itemize}

%\section{Honors and Awards}
%\textbf{University of Washington} \\
%\begin{tabular}{l l l}
%2022 & Z.W. Birnbaum Award (for best general exam) & \emph{Department of Statistics} \\
%2020 & Dorothy M. Gilford Excellence in Teaching Award & \emph{Department of Statistics}
%\end{tabular}  \\
%\\
%\textbf{Williams College} \\
%\begin{tabular}{l l }
%2018 & W. Marriott Canby Athletic Scholarship Prize   \\
%& (for highest standing in scholarship among senior varsity athletes) \\
%2018 & Robert M. Kozelka Prize in Statistics   \\
%& (presented to outstanding senior) \\
%2018 &  Sigma Xi Scientific Honor Society \\
%%5 nate kristof award would be great but no one will know what it is
%2017 & New England Small Colleges Athletic Conference (NESCAC) All-Sportsmanship award \\
%2017 & Phi Beta Kappa  National Honor Society \\
%2016-2018 & NESCAC All-Academic team \\
%2016-2018 & Clare Boothe Luce Fellowship \\
%& (Funded summer research and \$3,000 of research and travel funds) \\
%2016 & Erastus C. Benedict First Prize in Mathematics \\
%& (presented to outstanding sophomore) \\
%\end{tabular}
%
%
\section{Professional Service}
\textbf{Manuscript reviewer:} \textit{Annals of Statistics}, \textit{Journal of the American Statistical Association}, \textit{Annals of Applied Statistics}, \textit{Journal of Machine Learning Research}, \textit{Cell Systems}, \textit{Statistical Science}, \textit{Journal of Computational and Graphical Statistics}, \textit{Japanese Journal of Statistics and Data Science}, \textit{Electronic Journal of Statistics}, \textit{Biometrics}, \textit{Bernoulli}, and \textit{Genetic Epidemiology}. \\
\\
\textbf{Invited panelist:}
\begin{itemize}
\item Panel on navigating the academic job market, hosted by the Institute for Mathematical Statistics (IMS) New Researchers Group, November 2024, virtual. 
\item Panel on statistics graduate school, hosted by the Consortium for the Advancement of Undergraduate Statistics Education (CAUSE) at the  Electronic Undergraduate Statistics Research Conference, November 2022, virtual.
\end{itemize}



\section{Department or University \\ Service}
\textbf{Williams College}
\begin{itemize}
\item Statistics hiring committee, Fall 2024. 	
\end{itemize}
\textbf{University of Washington}
\begin{itemize}
\item Lead consultant and volunteer recruitment coordinator for UW StatCom (statistics in the community non-profit consulting program), 2022-2023. 
\item Member of UW Statistics Undergraduate Curriculum Committee, 2021-2022. 
 \item UW Statistics Graduate Student Representative (elected and paid position that involves attending faculty meetings, serving on committees, and planning orientation for new students). 2020-2021.
 \item Student interviewer for tenure track assistant professor search (2020) and teaching professor search (2021). 
\item First round reviewer for PhD Admissions, 2020-2023. 
\item UW Statistics Diversity, Inclusion, Community, and Equity (DICE) Committee, 2019-2023.   
\item Chair of the UW Statistics Fun Committee, 2019-2020.
  \item Organizer and Founder: \href{https://pearce790.github.io/statedrg.html}{Statistics Education Reading Group}, 2019-2020.
 \end{itemize}
%\textbf{Williams College}
% \begin{itemize}
%\item Member of the Committee on Undergraduate Life (committee consisted of students, faculty, and staff), 2015-2016.
%\item Member of the Great Ideas Committee (managed a budget of \$10,000 per year to implement improvements to undergraduate life), 2015-2017.
%\end{itemize}
 
\section{Outreach}
\begin{itemize}
   \item Lesson leader at \href{https://math.washington.edu/events/2023-03-20/uw-math-day}{UW Math Day}, March, 2023. 
  \item Lesson leader at \href{https://depts.washington.edu/gearupac/}{UW GEAR UP's} (Gaining Early Awareness and Readiness of Undergraduate Education) after school program (May 2021) and statistics summer camp (July 2022). 
\end{itemize}
%
%\section{Other \\
%Experience}
%{\textbf{Cogo Labs, Cambridge, MA}} \hfill June 2017-August 2017 \\
%Data Analytics Intern
%\begin{itemize}
%\item Worked with a team of engineers, designers, and analysts to build and market a website. Analyzed market data with SQL, analyzed site performance with google analytics and piwik, and assisted with backend web development in python. 
%\end{itemize}
%{\textbf{SMALL Research Experience for Undergraduates} } \hfill June 2016-August 2016 \\
%Williams College
%\begin{itemize}
%\item Advisors: Dr. Julie Blackwood and Dr. Lauren Childs
%\item Worked with a team of undergraduates in applied mathematics to model the dynamics of Zika Virus. Work resulted in co-first-author publication. 
%\end{itemize}
\end{resume}
\end{document}